---
title: "Assignment 1 - MA710 Bentley"
output: html_document
---

## Introduction

The data we are working on is provided by the U.S department of Education. The dataset contains information of College Scorecard, federal financial aid and students'earning records. The data set can help to understand the information of getting federal financial ai and have insights on the school performance as well as students' long term development.The dataset contains 1743 variables which are organized into 10 categories. The categories cover a range of topics:

*Root - basic information such as School Id and location.
*School - basic descriptive information such as degree types, number of branches, Carnegie classification, Public or Private etc.
*Academics - types of academics available at the school.
*Admissions - admission statistics such as acceptance rate and SAT/ACT scores.
*Student - descriptions of the students including the number of undergraduate students, part-time or full-time students, share of first generation students.
*Cost - information  cost for students such as average cost of attendance, tuiton and fees, and avergae net price for institutions by income level.
*Aid - the amount of debt a student can take on while attending the college, typical monthly loan payments, and the percentage of Pell Grant Recipients
*Repayment - descriptions of how successful students are at reapying their debts which is measured by cohort default rate and repayment rate.
*Completion - information on the compeletion and retention rates for first time and full time students
*Earnings - statistics on average and median earnings as well as share of former students' earning more than $25,000. 

Based on the information provided by the dataset,the descriptive analysis is performed. 


##Goals
"private colleges vs. public colleges"- there is a face-off that's been ongoing for many years. The major difference between public and private colleges lies in how they are funed. Public colleges are funded mostly by state governments, whereas private colleges rely heavily on private contributions, donations and tuition. The goal of our analysis is to compare public institutions and private colleges in terms of their **performance**.  A college degree or postsecondary certificate is more important than ever. However, many students are not confident enough to making decision on the type of school they should choose. Our analysis aims at helping students understand the differences between public institutions and private colleges in terms of school size, school performance, tuition cost, degrees awarded, students' background and geographic information. The analysis should give potential students a brief overview of these two types of schools. 


##DataSet Description

There are 26 variables that we are interested in. The variables name,  type and descriptions are listed below. 

*For variable NPT4_PRIV and NPT4_PUB, they includes tuiton, supplies, living expenses minus the average grand or scholarship; the cost is calculated for full-time, fist-time, degree and certificate seeking undergraduates

*For variable CCBASIC, the Carnegie Classification describes the colleges as either Associate's collegeSpecial focus, Baccalaureatte/Associate, Doctoral University, or Tribal college. 


|Variables        |Type     |Descriptions|
|-----------------|---------|-------------------------------
|Type             |FACTOR   |Public or Private Institution|
|State            |FACTOR   |State|
|Family_Inc       |FLOAT    |Average family income in real 2015 dollars|
|Female           |FLOAT    |Percentage of female students|
|Married          |FLOAT    |Percentage of married students|
|Ind_Per          |FLOAT    |Percentage of financially independent students|
|First_Gen        |FLOAT    |Percentage of first generation students|
|Degree           |FACTOR   |Predominant degree awarded |
|HIGHDEG          |FACTOR   |Highest degree awarded (0-non degree awarded;4-graduate degree)|
|REGION           |FACTOR   |The region where the college is located (10 levels)|
|NPT4_PRIV        |FLOAT    |Average annual cost of attendance a Private college|
|NPT4_PUB         |FLOAT    |Average annual cost of attendance a Public college|
|NUMBRANCH        |FLOAT    |Number of Branch Campuses|
|CCBASIC          |INTEGER  |Carnegie Classification which describes the colleges|
|PCIP01           |FLOAT    |Percentage of degrees awarded in Agriculture , Agriculture Operations, And Related Sciences|
|PCIP03           |FLOAT    |Percentage of degrees awarded in Natural Resources And Conservation|
|PCIP04           |FLOAT    |Percentage of degrees awarded in Architecture And Related Services|
|PCIP05           |FLOAT    |Percentage of degrees awarded in Area, Ethnic, Cultural, Gender, And Group Studies|
|PCIP09           |FLOAT    |Percentage of degrees awarded in Communication, Journalism, And Related Programs|
|PCIP10           |FLOAT    |Percentage of degrees awarded in Communications Technologies/Technicians And Support Services|
|CDR3             |FLOAT    |Three-year cohort default rate|
|CDR3_DENOM       |FLOAT    |Number of students in the cohort for the three-year cohort default rate|
|PCTPELL          |FLOAT    |Percentage of undergraduates who receive a Pell Grant|
|PCTFLOAN         |FLOAT    |Percent of all federal undergraduate students receiving a federal student loan|
|DEBT_N           |FLOAT    |The number of students in the median debt completers cohort|
|CUML_DEBT_N      |FLOAT    |Number of students in the cumulative loan debt cohort|



##Objectives
To better understand the differences between private colleges and public institutions, the descriptive statistical analysis is conducted. The detailed steps are describled as following:
1. Import and prepare dataset. 
This step imports dataset from local dictionary into R and read the file as csv file. All the necessary packages are installed in the R for later analysis and data visualization. Since we are interested in 26 variables, we select 26 variables from original dataset and create a new dataset for further analysis. The data manipulation is not implemented in the original dataset. 

2. Data preparation. 
In this step, in the new dataset we created, missing values for each variables are ensured less than 20% of the total values; data type is transferred into the proper type for analyzing convenience; the target variable Type, Private or Public is code as 0 for private and 1 for public; some independent variables are renamed for interpretation purpose. This step cleans and prepares all the data for further analysis. Dplyr is widely used in this step. 

3. Data summary
In this step, basic statistical analysis is conducted. The mean, median, and the variance is calculated for each numeric variables. This can help with the general understanding of the dataset. Dplyr is widely used in this step. 

4. Relationship among variables
In this step, the relationship between target variables and independent variables as well as the relationship among independent variables is analyzed. When it comes to the relationship among independent variables, the correlation among independent variables from the same category is tested as well as the correlation among independent variables cross different categories. This will help to get the insights on the difference between private colleges and public institutions. This step is crucial in achieving our goal. Ggplot is widely used in this step. 

5. Conclusion
This step summarizes the descriptive results for private colleges and public institutions. The descriptive analysis indicates the difference between private school and public school from perspectives such as students’ background, degree awarded, tuitions and stc. The result can help students understand private and public school better and help them make decision for their education goal. 
##Data Preparation

### 5.1 Libraries used
#### 5.1.1 "dplyr" package
We will load the *'dplyr'* package that contains functions for data manipulation using data frames. It allows us to order rows, select rows, select variables, modify variables and summarize variables. 
```{r}
library(dplyr)
```

#### 5.1.2 "ggplot2" package
We will load the *"ggplot2"* package which is a powerful plotting system that creates elegant and complex plots in R.
```{r}
library(ggplot2)
```
#### 5.1.3 "ggvis" package
We will also use the graphics commands from the ggvis library.
```{r}
library(ggvis)
```
#### 5.1.4 "magrittr" package
The magrittr library is loaded for the piping operator %>%.
```{r}
library(magrittr)
```


```{r}
# Read in the csv file using `read_csv` function
csb.df = read_csv(paste0('/Users/akhoury/Desktop/MA 710/Assignment-1/CollegeScorecard_Raw_Data/', 
                         'MERGED2014_15_PP.csv'))
```

```{r}
# List the variable names from the dataset
names(csb.df)
```

```{r}
# Remove those institutions(rows) which are not currently operating
csb.df %>%
  filter(CURROPER ==1) %>%
  {.} -> csbop.df
```

```{r}
# Look for variables to analyse and examine their class
csbop.df %>%
  select(CONTROL, ST_FIPS, FAMINC, FEMALE, MARRIED, 
          DEP_STAT_PCT_IND, FIRST_GEN, PREDDEG ) %>%
  {.} -> csb.vars.df
str(csb.vars.df)
```

```{r}
# Change the class of certain variables as needed
csb.vars.df %>% 
  mutate(FAMINC  = as.numeric(FAMINC),
         FEMALE = as.numeric(FEMALE), 
         MARRIED  = as.numeric(MARRIED),
         DEP_STAT_PCT_IND = as.numeric(DEP_STAT_PCT_IND),
         FIRST_GEN = as.numeric(FIRST_GEN),
         CONTROL    = as.factor(CONTROL),
         ST_FIPS = as.factor(ST_FIPS),
         PREDDEG = as.factor(PREDDEG)) %>%
         {.} -> csb.class.df
```


```{r}
# Rename the chosen variables
csb.new.df %>%
  rename(Type = CONTROL,
         State = ST_FIPS,
         Distance = DISTANCEONLY,
         Family_Inc = FAMINC,
         Female = FEMALE,
         Married = MARRIED,
         Ind_Per = DEP_STAT_PCT_IND,
         First_Gen = FIRST_GEN,
         Degree = PREDDEG) %>%
         {.} -> csb.ren.df
```


We can see that the selected variables are mostly without missing values. The exception is CCBASIC which has 421 missing values or the equivalent of 5.46% from the data. Since this is an important variable for us and considering the fact approximately 5% are msising values, we will use the CCBASIC variabe for our analysis.

##Variable Summaries

```{r}
# Check for the percentage of NA (missing) values for the above range of variables
csb.class.df %>% 
  summarise_at(.cols = vars(),
               .funs = function(x) sum(is.na(x))/length(x))*100
```

```{r}
# Select those variables with less than 20% missing values 
csb.class.df %>%
  select(CONTROL, ST_FIPS, DISTANCEONLY, FAMINC, FEMALE, MARRIED, 
         DEP_STAT_PCT_IND, FIRST_GEN, PREDDEG ) %>% 
         {.} -> csb.var.df
```

```{r}
# Combine private for-profit and private non-profit levels in the Type variable. Type = 1 for public,
# Type = 0 for Private
csb.var.df %>%
  mutate(CONTROL = ifelse((CONTROL==2|CONTROL==3),0,1)) %>%
  mutate(CONTROL    = as.factor(CONTROL)) %>%
  {.} ->csb.new.df
```



###Sev
We will use some graphs in order to explore further the variables and the relationship between them. Since we are considering the variables through the Public vs Private category, when exploring the variables we will use different colors to emphasize the differences which come with Private vs Public colleges.

(all of my part is moved into the next section; I did some summary of the data but that's basically the same as the rest of the team)

### Neha
## 5.1 Family Income, Female, Married, Percentage of financially independent student and Percentage of first generation students

The summary statistics for all the numeric variables were obtained using the `summarise_if` function. This function was used in combination with `is.numeric` to summarise only those variables which were numeric. We also grouped these variables by their `Type` which is Public or Private, beofre summarising.The descriptive measures chosen were mean, median and variance. The `options` command was used to set the column width to infinity inorder to see all the columns.The code used was : 
```{r}
options(dplyr.width=Inf)
csb.ren.df %>% 
  group_by (Type) %>%
  summarise_if(is.numeric, funs(mean, median, var), na.rm=TRUE)
print(Rail)
```
We can observe how these measures vary considerably for public and private institutions.Though the average family income for both type of institutions is roughly similar, the median family income for Public insitutions ($ 31,000) is ~$3,000 more than that of Private institutions. This is an interesting finding as private institutions being more expensive, should be expected to be attended by students coming from the higher income bracket. It can also be seen that private insitutions have a larger share of female students, married students, financially independent and first-generation students.

## 5.2 Degree and type

The following code is used to summarise the factor variables usig aggregates. We first group the institutions by their type and count the total number of institutions providing predominantly Certficate, Associate, Bachelor and Graduate degrees. The code used is:
```{r}
csb.ren.df %>% 
  group_by(Type) %>%
  summarise(Certificate=sum(Degree==1),
            Associate=sum(Degree==2),
            Bachelor=sum(Degree==3),
            Graduate=sum(Degree==4))
```
The output tells us that there are 2904 institutions providing predominantly Certificate degree, 1298 institutions with Bachelor and 285 Graduate degree institutions.

###Cheng

To get a better understanding of our variables, we could adopt descriptive analysis to summary independent varibles. Furthermore, we could generate summary statistics respectively for two different categories of our target varivble- CONTROL:
```{r}
options(dplyr.width=Inf)
csb.vars.df %>% 
  group_by (CONTROL) %>%
  summarise_if(is.numeric, funs(mean, median, var), na.rm=TRUE)
```

From the output, we could summarise that in terms of mean value:


for variables *PCTPELL*, *PCTFLOAN*, *DEBT_N*, *CUML_DEBT_N*, public institutions(CONTROL=1) are much higher than Private insitituions.

for variables *CDR3*, *CDR3_DENOM*, public institutions(CONTROL=1) are much lower than Private insitituions.


In terms of median value:

for variables *PCTPELL*, *PCTFLOAN*, *CUML_DEBT_N*, public institutions(CONTROL=1) are higher than Private insitituions.

for variables *DEBT_N*, *CDR3*, *CDR3_DENOM*, public institutions(CONTROL=1) are lower than Private insitituions.


In terms of variance:

for variables *PCTPELL*, , *CUML_DEBT_N*, *DEBT_N*, *CDR3*, *CDR3_DENOM*, public institutions(CONTROL=1) are higher than Private insitituions.

for variable *PCTFLOAN* public institutions(CONTROL=1) are slightly lower than Private insitituions.

In conclusion, the summary statistics reveals that considerably differnece exists between two catories of target varible. 

###Katie
```{r}
options(dplyr.width=Inf)
csb.vars.df %>% 
  group_by (CONTROL) %>%
  summarise_if(is.numeric, funs(mean, median, var), na.rm=TRUE)
```
The result indicated that the averages of the percentage of awarded degree in those 6 fields are less than 1% .Among those 6 awarded degree, the mean of the PCIP09 is the largest regardless of the contorl type.This result indicates that among those 6 variables, percentages of degrees awarded in Communication are awarded more than otehr degree.The variance of those 6 variables are relatively samll, which means that from 2014 to 2015,the percentage of those 6 awarded degrees are almost the same in general. It does not vary to much in different institutes.PCIP01,PCIP03,PCIP05,PCIP09 degrees are awared more in public school than privated school. 





##Variable relationship

###Sev
In this section we will focus on comparing the Public vs Private colleges based on some basic characteristics such as location by region, level of degrees offered, number of campuses etc.

```{r, cache=TRUE}

#Total number pf Public vs Private colleges = there are at least twice as more Private colleges compared to Public
merged2014.vars.df %>%
  ggplot  (aes(x   = as.factor(CONTROL))) +
  geom_bar(aes(fill=CONTROL)) 
```

From the graph above we can see that there are appriximately 2,000 Public colleges in the data set we are exploring. The number of Private colleges is more than twice the number of Public colleges.That means that we will have more data available for Private colleges and we will be able to draw more conclusion for them based on the dataset.


The Region variable gives information which region a college belongs to. The graph below shows the distribution of colleges by region.

```{r, cache=TRUE}
#College count by Region colored by Public vs Private
merged2014.vars.df %>%
  ggplot(aes(x=as.factor(REGION))) +
  geom_bar(aes(fill= as.factor(CONTROL)))
```

Based on the graph, we can see that the colleges are not equally distributed among regions. Some regions such as numbers 2(Mid East),3(Great Lakes) and 5(Southeast) have much  larger amount of colleges compared to 7(Rocky Mountains) and 9(Outlying areas). In addition the ratio of the Public vs Private colleges changes between regions.

The variable CCBASIC describes the Carnegie classification of colleges. There are 33 different type of colleges ranging between Associate's college and Doctoral Universities. We will explore the distribution of colleges based on the CCBASIC values.

```{r, cache=TRUE}

#College count by CCBASIC colored by Public vs Private
#
merged2014.vars.df %>%
  ggplot(aes(x=as.factor(CCBASIC))) +
  geom_bar(aes(fill= as.factor(CONTROL)),
            position='dodge')

```
From the plot, we can see that some categories are stronger represented. A significant amount fo the values belong to category (-2) which means that the classification is not applicable. In addition we can see that Public colleges are represented better in some categories(1,2,3,4,5) while private colleges are strongly represented in other categories such as 21,22,23. In order to be able to explore the graph in more details, we will remove the nor classified colleges and will split the rest of the data in two parts - csategories which are strognly represented by Public colleges and these which are strongly represented by Private colleges.


```{r, cache=TRUE}
merged2014.vars.df %>% 
  filter(CCBASIC %in% c(1:9, 14:18)) %>%
  {.} -> merged2014.vars.pub.ccbasic.df

merged2014.vars.df %>% 
  filter(CCBASIC %in% c(10:13, 19:33)) %>%
  {.} -> merged2014.vars.priv.ccbasic.df

merged2014.vars.pub.ccbasic.df %>%
  ggplot(aes(x=as.factor(CCBASIC))) +
  geom_bar(aes(fill= as.factor(CONTROL)),
            position='dodge')

merged2014.vars.priv.ccbasic.df %>%
  ggplot(aes(x=as.factor(CCBASIC))) +
  geom_bar(aes(fill= as.factor(CONTROL)),
            position='dodge')

```

From the new plots, we can see that 9 (Associate's Colleges: High Vocational & Technical-High Nontraditional) and  14(Baccalaureate/Associate's Colleges: Associate's Dominant) are with almost equal numbers of Public and Private colleges. The rest of the categories are strognly dominated by one type or another. It's interesting to see that most Associates' colleges are mostly Public with very few Private options. On the other hand the Special focus Two-Years and the Baccalaureatte colleges are mostly Private ones.

The next variable to explore is the Number of Branches. Again we will see the distribution colored based on the Public vs Private options.

```{r}
#College count by NUMBRANCH colored by Public vs Private
#
merged2014.vars.df %>%
  ggplot(aes(x=as.factor(NUMBRANCH))) +
  geom_bar(aes(fill= as.factor(CONTROL)),
            position='dodge')
```

From the graph, it is visible that the Private colleges are the ones that offer more branches. The Public colleges rarely has more than 6 branches. The majority of the colleges are with only one branch.

The next variable in our exploratory analysis is the HIGHDEG. It describes the highest degree awarded at the college. 

```{r, cache=TRUE}
merged2014.vars.df %>%
  ggplot(aes(x=as.factor(HIGHDEG))) +
  geom_bar(aes(fill= as.factor(CONTROL)),
            position='dodge')
```

From the graph we can see that categories 0 (no degree), 1(Certificate) and 3(Bachelors) are very strognly represented by Private colleges. The Public colleges are stronger is giving Associate degrees and Gradyate degrees.

The last variable to explore is the Tuiton(NPT4).


```{r, cache=TRUE}
#College count by NPT colored by Public vs Private
merged2014.vars.df %>%
  ggplot(aes(x=as.factor(CONTROL),
             y=NPT4)) + 
  geom_boxplot(aes(fill=CONTROL)) 
```


Based on the box plots, we can see that the Private colleges(on the left) are more expensive compared to Public with some very expensive private colleges presented as outliers on the plot. The median of the tuiton for Public colleges is also lower than the median of the tuiton for Private colleges. The cheaper Private colleges are with comparable tuiton to the most expensive Public colleges.

The next step is to explore some variable relationships.

The first relationship to focus on is whether there is a strong relationship between the tuiton and the number of branches.

```{r}
#NPT4 by Number of branches
ggplot(data=merged2014.vars.df) + 
  aes(x=NPT4,
      y=NUMBRANCH) + 
  geom_point(size=1,
             alpha=0.2)

```

Based on the plot, we can see that there is no clear correlation. The majority of the points are at the bottom of the graph which is explained by the fact that most of the colleges has 1 branch only. We can see that some colleges with very high number of branches, which stands out on the top of the graph, are with medium tuiton in the range of 20,000 - 25,000. And the most expensive colleges has 1 branch only. Therefore we can not find a clear correlation between tuiton and number of branches.

The next relationship, we are interested in is to see how the tuiton relates to region. The question we are exploring is whether some regions are more expensive than other.

```{r}
#NPT by REGION and by Public
merged2014.vars.df %>%
  ggplot(aes(x=as.factor(REGION), y=NPT4)) +
  geom_boxplot(aes(fill=as.factor(CONTROL)))

```

The graph represents the tuiton by region and by Public vs Private colleges. We can see that the tuiton for the Public colleges is in general lower for each of the 9th regions. We can also observe that most of the regions are with comparable tuition when we compare the Private colleges and the Public colleges separately. Region 1(New England) stands out as the most expensive region when it comes to both Public and Private. While region 9 (Outlying areas) stands out as the cheapest area for education. Region 5(Southeast) has some very expensive colleges which stands out as outliers.

The next comparison is between the Classification and the Tuiton. We are exploring whether the tuiton varies by the type of college.

```{r}
#NPT by CCASIC and by Public
merged2014.vars.df %>%
  ggplot(aes(x=CCBASIC, y=NPT4)) +
  geom_point(aes(color=CONTROL))

```

As we already discovered in previous graphs, certain type of colleges such as Associates are more strongly represented by Public colleges. As a result their tuiton is lower than the tuiton of other categories which are strongly represented by Private colleges such as Special focus Two Year colleges.

It is also interesting to explore whether some regions tend to have colleges with more branches and how that is affected by Public vs Private classification.


```{r}
#NUMBRANCH by REGION and by 
merged2014.vars.df %>%
  ggplot(aes(x=as.factor(REGION), y=NUMBRANCH)) +
  geom_point(aes(color=CONTROL))

```

From the graph, we can see that the number of branches are approximately equally distributed between the regions. Region 1(New England) and Region 9 (Outlying areas) have colleges which are in general with slightly less number of branches. That could be explained by the geographical size of the resions. If the region is larger then it would make sense for the college to expand the number of branches in order to cover larger students' population.

###Neha
## 6.1 Financially Independent, First Generation and Family Income

In order to visualise the relationship between the three numeric variables, family income, precentage of students who are financially independent and the percentage of first generation students, we use a tile graph. The x-axis represents the Financially independent students in percentage while the y-axis shows the percentage of first generation students. The color of the tile depicts the family income of students, white tiles for the low values and red for the high values of family income. The code used is :
```{r}
csb.ren.df %>%
  ggplot +
  aes(x=Ind_Per, 
      y= First_Gen, 
      z= Family_Inc) +
  stat_summary2d(fun=mean) +
  scale_fill_gradient(low="white", 
                      high="red") 
```
It can be observed from the graph that the there is a strong positive correlation between first generation students and financially independent students. This implies that institutes that have a higher share of first generation students also seem to have a higher share of financially independent students. Another observation is that the red tiles are at the lower end of both axes, indicating that institutes with students coming from a background of higher family income tend to have lower percentage of both first generation and financially independent students. 

## 6.2 Family Income, Degree and Type

We now try to analyse the relationsip between family income of students, degree provided by institutons and the type of the insitution.Since it involved one numeric and two factor variables, we use a heat map.The x-axis represents the type, the y-axis denotes the degree and the color of the boxes stands for the average family income of students.The code used is:

```{r}
csb.ren.df %>%
  group_by (Type, Degree) %>%
  summarize(avg.inc=mean(Family_Inc, na.rm = TRUE)) %>%
  ggplot   (aes(x=Type, y=Degree)) + 
  geom_tile(aes(fill=avg.inc)) 
```

This output clearly depics that institutes with students having the highest average famnily income predominantly award Bachelors degree (Degree = 3). This is true for both public and private institutions. On the other hand, institutes that award Certificate degrees have students with the lowest family income. This could partly be explained by the high cost of the 4-year Bachelors program.

## 6.3

The code below is used to count the number of colleges in each state and arrange them from highest to lowest.
```{r}
csb.ren.df %>%
  group_by(State) %>%
  summarise(Num_of_Public = sum(Type == 1),
            Num_of_Private = sum(Type == 0)) %>%
  arrange(desc(Num_of_Public+Num_of_Private)) %>%
  print (n=3)
```

From the output it can be seen that States 6,36 and 48 have the highest total number of colleges. These states are California, New York and Texas.For these top 3 states, we now would like to know which are the most popular degrees.

```{r}
csb.ren.df %>%
  filter(State %in% c(6,36,48)) %>%
  ggplot(aes(x=State)) + 
  geom_bar(aes(fill=Degree), 
           position='dodge') 
```

Certificate degrees seem to be the most popular in State 6 and 48 which are California and Texas. It is the second most popular degree after Bachelor's degree in New York. For these most popular certificate degrees, we wish to find the relationship between the type of institution, first generation students % and the family income of students.

```{r}
csb.ren.df %>%
  filter(State %in% c(6,48,36) & Degree ==1) %>%
  ggplot(aes(x=Family_Inc, y=First_Gen)) +
  geom_point(aes(color=Type, size=3, alpha= 0.2)) + 
  guides(size=FALSE)
```

We observe a negative correlation between the family income of students and the percentage of first generation students. This tell sus that for the most popular Certificate degrees in the most poular states California, New York and Texas, first generation students have lower family income. 

###Cheng
### 7.1 Target variable 'CONTROL'
We use a bar plot to compare numbers of observations for the target variable *CONTROL*
```{r}
csb.vars.df %>%
  ggplot  (aes(x   =CONTROL)) +
  geom_bar(aes(fill=CONTROL))
```

From the plot above we can find that there are appriximately 2,000 Public insitituions. The number of Private colleges is nearly 5700.

### 7.2 Target variable ~ independent variables
We use box plot to compare distributions of each independent variable between two levels of target variable.
### 7.2.1 "CONTROL"~"CDR3"
```{r}
csb.vars.df %>%
  ggplot(aes(x=CONTROL,
             y=CDR3)) + 
  geom_boxplot(aes(fill=CONTROL)) +
    ylim(0, 0.6)
```

### 7.2.2 "CONTROL"~"CDR3_DENOM"
```{r}
csb.vars.df %>%
  ggplot(aes(x=CONTROL,
             y=CDR3_DENOM)) + 
  geom_boxplot(aes(fill=CONTROL)) +
  ylim(0, 10000)
```

### 7.2.3 "CONTROL"~"PCTPELL"
```{r}
csb.vars.df %>%
  ggplot(aes(x=CONTROL,
             y=PCTPELL)) + 
  geom_boxplot(aes(fill=CONTROL))
```

### 7.2.4 "CONTROL"~"PCTFLOAN"
```{r}
csb.vars.df %>%
  ggplot(aes(x=CONTROL,
             y=PCTFLOAN)) + 
  geom_boxplot(aes(fill=CONTROL))
```

### 7.2.5 "CONTROL"~"DEBT_N"
```{r}
csb.vars.df %>%
  ggplot(aes(x=CONTROL,
             y=DEBT_N)) + 
  geom_boxplot(aes(fill=CONTROL)) +
  ylim(0, 30000)
```

### 7.2.6 "CONTROL"~"CUML_DEBT_N"
```{r}
csb.vars.df %>%
  ggplot(aes(x=CONTROL,
             y=CUML_DEBT_N)) + 
  geom_boxplot(aes(fill=CONTROL)) +
  ylim(0, 10000)
```

From all the box-plot above, we could conclude that the distribution of each variable are quite different for two levels of the target varible. We could draw the conclusion and achieve our initial goal based on our analysis output.


###Katie

#Target variables~independent variables:
Control~PCIP01:
```{r}
csb.vars.df %>%
  ggplot(aes(x=CONTROL,
             y=PCIP01)) + 
  geom_boxplot(aes(fill=CONTROL)) +
  ylim(0, 0.6)
```
The result indicates that the major percentage of taking argriculture degree is less than 0.1 percent in private school and less than 0.2 percent in public school. There are more outlier in private school. 
```{r}
csb.vars.df %>%
  ggplot(aes(x=CONTROL,
             y=PCIP03)) + 
  geom_boxplot(aes(fill=CONTROL)) +
  ylim(0, 0.7)
```
The result indicates that the majority percentage of taking natural resource degree is less than 0.15 percent in public school and less than 0.1 percent in private school.There are more outliers in private school than in public school.In some private school, the natural degree is awarded more than other private school and public school.
```{r}
csb.vars.df %>%
  ggplot(aes(x=CONTROL,
             y=PCIP04)) + 
  geom_boxplot(aes(fill=CONTROL)) +
  ylim(0, 0.5)
```
The result indicates that the majority percentage of taking Architecture degree is less than 0.07 percent in private and public school. Private school has higher Architecture degree pecentage than public school.There are more outliers in private school.
```{r}
csb.vars.df %>%
ggplot(aes(x=CONTROL,
           y=PCIP05)) + 
  geom_boxplot(aes(fill=CONTROL)) +
  ylim(0, 0.5)
```
The result indicates that the majority percentage of Ethnic degree in private school is less than 0.1 percent while it is less than 0.05 in public school. There more higher percentage of Ethnic dergree awarded in public school because public school has more outliers.
```{r}
csb.vars.df %>%
  ggplot(aes(x=CONTROL,
             y=PCIP09)) + 
  geom_boxplot(aes(fill=CONTROL)) +
  ylim(0, 0.6)
```
The result indicates that majority of percentage of communication/juornal degree awarded is less than 0.025 in public school.Private school has more high percentage communication degress awarded and has more outliers.
```{r}
csb.vars.df %>%
  ggplot(aes(x=CONTROL,
             y=PCIP10)) + 
  geom_boxplot(aes(fill=CONTROL)) +
  ylim(0, 0.5)
```
The result indicates that the range of percentage of communication technology degree awarded in private school is larger than public school. Private school has higher percentage of communication technology degree awarded.The percentage of communication techonology degree awarded is more spread in private school than in public school. 


#Among variables
```{r}
cor_matrix<-csb.vars.df %>%
  group_by(CONTROL)%>%
  do(as.data.frame(cor(.[,-1],method="spearman",use="pairwise.complete.obs")))
cor_matrix1<- cor_matrix%>%
  data.frame(row=rep(colnames(.)[-1],n_groups(.)))
cor_matrix1
```
Among private school, PCIP03 and PCIP05 has some correlation, around 0.53.Among public school, PCIP05 and PCIP09 has some correlation around 0.46. 

##Conclusion




